---
title: 'STA426: Exercise 2'
author: "Balint Tamasi"
date: "9/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/btamasi/UZH-biostat/2017FS/STA426 - High Throughput Genomic/material/week02-25sep2017/exercise")
```

# Data Import

```{r}
anno = read.table("SampleAnnotation.txt", as.is=TRUE, sep="\t", quote="",
                  row.names=1, header=TRUE)
x = read.table("expressiondata.txt", as.is=TRUE, sep="\t", quote="", row.names=1, header=TRUE, check.names = FALSE)
x = as.matrix(x)
```

## Define samples and colors and phenotype
```{r}
samples = rownames(anno)
colors = rainbow(nrow(anno))
isNorm = anno$TissueType == "norm"
isSick = anno$TissueType == "sick"
isAcute = anno$TissueType == "acute"
```

# Distributions

## Boxplot

```{r}
boxplot(log(x, 2), col = colors)
```

## Density

A single density plot

```{r}
plot(density(log(x[, 1], 2)))
```

## Limma::plotDensities

```{r}
limma::plotDensities(log(x, 2), legend = "topright", col = colors)
```

# Normalization

```{r}
x_norm <- limma::normalizeQuantiles(x)
limma::plotDensities(log(x_norm, 2), legend = "topright", col = colors)
```

# Clustering

Using `1-corr(x)` as distance measure.

First, let's do it column-wise:

```{r}
dis_col <- as.dist(1 - cor(x_norm))
cl_col <- hclust(dis_col)
plot(cl_col)
```

Secondly, do it row-wise for 1000 randomly selected rows.

```{r}
set.seed(13)
x_norm_smpl <- x_norm[sample(nrow(x_norm), 1000), ] ## sample the normalized data
dis_row <- as.dist(1 - cor(t(x_norm_smpl)))
cl_row <- hclust(dis_row)
plot(cl_row)
```

# Heatmap

Using the `gplots::heatmap.2` function and the dendrograms computed in the previous steps.

```{r}
gplots::heatmap.2(x_norm_smpl, Rowv = as.dendrogram(cl_row), Colv = as.dendrogram(cl_col))
```

# Correlation matrix

```{r}
image(cor(x_norm), col = grey(seq(0, 1, length.out = 256)), axes = FALSE)
axis(1, at = seq(0, 1, length.out = ncol(x_norm)), labels = samples)
axis(2, at = seq(0, 1, length.out = ncol(x_norm)), labels = samples)
```

# Dimensionality reduction

## Multidimensional scaling

```{r}
mds <- cmdscale(dis_col)
plot(mds, col = ifelse(anno$TissueType == "norm", "green", 
                       ifelse(anno$TissueType == "sick", "orange", "red")),
     pch = 16)
legend("bottomleft", c("norm", "sick", "acute"), pch = 16, 
       col = c("green", "orange", "red"))
title("Multidimensional scaling")
```


## PCA

```{r}
pc <- prcomp(x_norm)
plot(pc)
summary(pc)
```

The first principal component accounts for almost 95% of the variance in the normalized data.